Tropical methods in Lambda-Calculus?

### 2023-03-16 rebuttal.txt

We thank the anonymous reviewers for their comments and suggestions.
We reply to a few of them below.

Reviewer 1:

> "which can be written as a parameterized power series of the form" what does this mean? $f$ is a single function, but $f_t$ is an infinite family of functions. does it mean $f$ can be written as $f_t$ for some fixed constant $t$?"

Rigirare la frittata su f, f_t ecc.

> Theorem 24 to bottom of column: I found this discussion hard to follow.

One of the core aspects of Theorem 24 is that on any L-module one can define a canonical metric; the goal of this paragraph is thus to describe in an explicit way the metrics associated with the SMCC structure of LMOD (or, equivalently, the metrics defining the SMCC structure of LCCAT). Yet we think that we should have noted X,Y rather as M,N, in order to avoid confusion.

Concerning tensors, in the previous page (after Prop. 19) we introduced the tensor product in LMOD and recalled that an element of it is a join of basic tensors x(x)y.

Reviewer 2:

> "Some of the calculations in the appendix regarding the differential story can be referenced away."

We thank the reviewer for his careful reading of the whole paper (including the long appendix), and for his numerous suggestions concerning results appearing in the recent literature. We were not aware of Lemay's "Coderelictions for Free Exponential Modalities" and we found its results very interesting. We will adjust our presentation following such remarks.

As a general remark, our results on LREL and LREL_! in Section III are based on relatively unsurprising checks (yet, it is always interesting to know that one can avoid long checks and use a theorem instead!). Since this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model, studied in more detail in the following sections.

> "Theorem 81 is important and should have appeared in the main paper."

Theorem 29 is (essentially) Theorem 81 from the appendix, but without mention of differential linear categories. We chose not to mention the latter in the main text (and thus to limit our discussion of differential operators to the cartesian case) because our results concerning them are either folklore or based on well-known results/techniques.

> "In the introduction, you say "the somehow canonical "Taylor semantics for STdLC". Why is that? What is the justification for this statement?"

...

> "I suggest rewriting Section III.D based on the comments made above, and explain that the differential structure comes from the fact that LREL is a differential category."

OK? Connettere alla discussione sopra??

Reviewer 3:

> "It is good that the model can characterize
    differential privacy of programs, but its implication to the
    verification tasks is not discussed."

Of course, the applications in Section 5 are just sketched. Our main goal in this paper was to demonstrate that tropical methods, which are widely applied in computer science (but not in the linear logic/lambda-calculus community) can make sense also for studying the quantitative semantics of the lambda-calculus. Concrete developments of such applications are thus left for future work.

> "Another limitation of the
    paper's model is that it can not handle probabilistic programs
    involving continuous distributions"

This is indeed a well-known limitations of the semantics of probabilistic programs based on weighed relational models (e.g. probabilistic coherent spaces [33]). Still, such semantics are well-studied and led to strong results (e.g. full abstraction theorems). Still, we agree that this limitation should be mentioned.

> "In complex analysis, Laurent series include terms of negative
    degrees, while here I think we only consider terms with
    non-negative degrees."

This is true, yet in the literature on tropical mathematics (to which we refer, e.g. [59]) these are defined only with positive coefficients. We agree that this should be explicitly mentioned, in order to avoid confusion.

> "Is Theorem 24, which is one of the main claims in the paper,
    original, or subsumed by [71]?"

We think that the main contribution in Section 6 is to show that an abstract correspondence known in the categorical literature can be extended to account for the relevant structure for interpreting the differential lambda-calculus. Hence we see Theorems 27 and 29, rather than Theorem 24 (whose statement and proof can be deduced from the results of [71]) as the main claims of this section.

### 2023-03-17 rebuttal.txt

We thank very much all the reviewers for the careful comments.
We will of course take care in correcting all the typos and unclear sentences.
We thank in particular reviewer 2 for his/her thorough reading, and we will perform the useful modifications he/her suggested (see below).

@@

Reviewer 1:

> p3 (1)

One just takes the semimodule generated by differential terms with coefficients in Q (or another semiring). Equivalently, you can see them as formal linear combinations with coefficients in the choosen semiring.

> p5 "setting $t^!(x) := t o_! x$"...

We are tacitly making some standard identifications: one identifies Q^X with the morphisms in QRel_!(T,X)=QRel(!T,X) where T=\emptyset is the terminal object. This is possible by mapping x \in Q^X to x' \in QRel(!T,X) defined by x'_{\emptyset,a} := x_a (remark that !T={\emptyset}). In this way, given t \in QRel_!(X,Y), one defines the function t^! : Q^X --> Q^Y by t(x) := t °_! x', which is a morphism T -> Y in QRel_!, seen in the same way as an element of Q^Y.

>  p7 what is $[[M]]^!$?

It is the function just defined in the previous answer (i.e. in equation (6) at page 5).

> p7 "which can be written as a parameterized power series of the form" what does this mean? $f$ is a single function, but $f_t$ is an infinite family of functions. does it mean $f$ can be written as $f_t$ for some fixed constant $t$?"

@@

Rigirare la frittata su f, f_t ecc.

> p9 Prop 17: why is the input of $f$ suddenly $[0, 1]^X$?

In the discussion around Def2 we were informal on what we meant by "database". Prop17 is, instead, a rigorous statement, so we stated it with the correct types for f. We are supposing to see a database as a function X --> [0,1], which is indeed often the case (for instance, for a relational database -- take X as the key/attributes and [0,1] as the datas).

@@

OK? Connettere alla discussione sopra??

Reviewer 3:

> "It is good that the model can characterize
    differential privacy of programs, but its implication to the
    verification tasks is not discussed."

You are right, the applications in Section 5 are just sketched. Our main goal in this paper was to demonstrate that tropical methods, which are widely applied in computer science (but not in the linear logic/lambda-calculus community) can make sense also for studying the quantitative semantics of the lambda-calculus. Concrete developments of such applications are thus left for future work.

@@

> "Another limitation of the
    paper's model is that it can not handle probabilistic programs
    involving continuous distributions"

This is indeed a well-known limitations of the semantics of probabilistic programs based on weighed relational models (e.g. probabilistic coherent spaces [33]). Still, such semantics are well-studied and led to strong results (e.g. full abstraction theorems). Still, we agree that this limitation should be mentioned.

> p.4 where is this hat operation defined?

It is supposed to be defined by those line. We agree that the sentence is not clear, so we will rephrase it.

@@

> "In complex analysis, Laurent series include terms of negative
    degrees, while here I think we only consider terms with
    non-negative degrees."

This is true, yet in the literature on tropical mathematics (to which we refer, e.g. [59]) these are defined only with positive coefficients. We will explicitly mention it, in order to avoid confusion.

> "Is Theorem 24, which is one of the main claims in the paper,
    original, or subsumed by [71]?"

We think that the main contribution in Section 6 is to show that an abstract correspondence (expressed indeed by Theorem 24) known in the categorical literature can be extended to account for the relevant structure for interpreting the differential lambda-calculus. Hence, we see Theorems 27 and 29, rather than Theorem 24 (whose statement and proof can be deduced from the results of [71]) as the main claims of this section.

### 2023-03-17 rebuttal.txt

We thank very much all the reviewers for the careful comments.
We will of course take care in correcting all the typos and unclear sentences.
We thank in particular reviewer 2 for his/her thorough reading (and numerous suggestions concerning results appearing in the recent literature), and we will perform the useful modifications he/her suggested (see below).

@@

> Theorem 24 to bottom of column: I found this discussion hard to follow.

One of the core aspects of Theorem 24 is that on any L-module one can define a canonical metric; the goal of this paragraph is thus to describe in an explicit way the metrics associated with the SMCC structure of LMOD (or, equivalently, the metrics defining the SMCC structure of LCCAT). Yet we think that we should have noted X,Y rather as M,N, in order to avoid confusion.  
Concerning tensors, in the previous page (after Prop. 19) we introduced the tensor product in LMOD and recalled that an element of it is a join of basic tensors x(x)y.

Reviewer 2:

> Comment 1 ("Some of the calculations in the appendix regarding the differential story can be referenced away"), comment 2, comment 7, comment 8.

We were not aware of Lemay's "Coderelictions for Free Exponential Modalities" and we found its results very interesting. We will surely adjust our presentation following such remarks! In particular, we will follow your suggestion to "carefully providing a proper reference for all the important math definitions".

As a general remark, our results on LREL and LREL_! in Section III are indeed based on relatively unsurprising checks; since this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model, studied in more detail in the following sections.
Therefore, our plan is to adjust Section III by making clearer what results are general (by also replacing the use of "Fact" with "Proposition/Lemma", toghether with the appropriate reference), and that the point of this section is to show how they look like in a tropical setting.
We would also accordingly adjust the appendix in case it should appear in an eventual final version of the paper.

@@

> "Theorem 81 is important and should have appeared in the main paper."

Theorem 29 is Theorem 81 from the appendix, except for the mention of differential linear categories. We chose not to mention the latter in the main text (and thus to limit our discussion of differential operators to the cartesian case) because our results concerning them are either folklore or based on well-known results/techniques.

@@

> "In the introduction, you say "the somehow canonical "Taylor semantics for STdLC". Why is that? What is the justification for this statement?"

We did not mean it to be a formal statement, but just the fact that it is known that Taylor approximation and relational semantics are tightly related, so it is natural to look at what happends in that kind of models. We agree that the sentence is not clear, so will modify it.

### 2023-03-18 rebuttal.txt

We thank very much all the reviewers for the careful comments.
We will of course take care in correcting all the typos and unclear sentences.
Below we anwer to the main points.

Reviewer 1:

> p5 "setting $t^!(x) := t o_! x$"...

We are tacitly using the standard indentification of Q^X with the morphisms in QRel_!(T,X)=QRel(!T,X) where T=\emptyset is the terminal object.

@@

>  p7 what is $[[M]]^!$?

It is the function discussed in the previous answer (i.e. in equation (6) at page 5).

> p7 "which can be written as a parameterized power series of the form" what does this mean? $f$ is a single function, but $f_t$ is an infinite family of functions. does it mean $f$ can be written as $f_t$ for some fixed constant $t$?"

We mean that we consider f a power series of shape \sum_n a_n x^n, and for fixed t>0 we take c_n := log_t (a_n), so that f can be written in the shape given in the paper. We agree that the subsection is not clear, so we will rephrase it better.

@@

> p9 Prop 17: why is the input of $f$ suddenly $[0, 1]^X$?

We are supposing to see a database as a function X --> [0,1], which is indeed often the case (for instance, for a relational database -- take X as the key/attributes and [0,1] as the datas). We will mention that.

> Theorem 24 to bottom of column: I found this discussion hard to follow.

One of the core aspects of Theorem 24 is that on any L-module one can define a canonical metric; the goal of this paragraph is thus to describe in an explicit way the metrics associated with the SMCC structure of LMOD (or, equivalently, the metrics defining the SMCC structure of LCCAT). Maybe we should use M,N rather than X,Y in order to avoid confusion.  
Concerning tensors, in the previous page (after Prop. 19) we introduced the tensor product in LMOD and recalled that an element of it is a join of basic tensors x(x)y.

@@

Reviewer 2:

First of all, we thank you for your thorough reading. We will surely adjust our presentation following your useful remarks!

> Comment 1, comment 2, comment 7, comment 8.

We were not aware of Lemay's "Coderelictions for Free Exponential Modalities" and we found its results very interesting/. As a general remark, our results on LREL and LREL_! in Section III are indeed based on relatively unsurprising checks; but since this paper is also addressed to the program metrics community, which is less familiar with linear logic and weighted relational models, this section is mainly thought to provide a relatively concrete presentation of (the tropical variant of) this model, before studying it in more detail in the following sections.
Therefore, our plan is to adjust Section III by making clearer what results are general (by also replacing the use of "Fact" with "Proposition/Lemma" toghether with the appropriate reference), and making clearer that the point of that section is to show how they look like in a tropical setting.
We would also accordingly adjust the appendix in case it should appear in an eventual final version of the paper.

@@

> "Theorem 81 is important and should have appeared in the main paper."

Theorem 81, from the appendix, appears in the main text as Theorem 29, except for the mention of differential linear categories. We chose not to mention the latter in the main text because our results concerning them are either folklore or based on well-known results/techniques.

@@

> "In the introduction, you say "the somehow canonical "Taylor semantics for STdLC". Why is that? What is the justification for this statement?"

We agree that the sentence is not clear, so will modify it. We simply meant that since Taylor and relational semantics are tightly related, it is natural to look at what happens in that kind of models.

@@

Reviewer 3:

> "It is good that the model can characterize
    differential privacy of programs, but its implication to the
    verification tasks is not discussed."

You are right, the applications in Section 5 are just sketched. However, one of the main goals of this paper is to demonstrate that tropical methods, which are widely applied in computer science (but not in the linear logic/lambda-calculus community) can make sense also for studying the quantitative semantics of the lambda-calculus. Concrete developments of such applications are thus left for future work. We will clarify that point.

@@

> "Another limitation of the
    paper's model is that it can not handle probabilistic programs
    involving continuous distributions"

This is indeed a well-known limitation of the semantics of probabilistic programs based on weighed relational models (e.g. probabilistic coherent spaces [33]). Still, such semantics are well-studied and led to strong results (e.g. full abstraction theorems). Anyway, we agree that this limitation should be mentioned.

@@

> p.4 where is this hat operation defined?

It is supposed to be defined by those line. But we agree that it is not clear, so we will rephrase it better.

@@

> "In complex analysis, Laurent series include terms of negative
    degrees, while here I think we only consider terms with
    non-negative degrees."

This is true, yet often in the literature on tropical mathematics (to which we refer, e.g. [59]) these are defined only with positive coefficients. We will explicitly mention it.

@@

> "Is Theorem 24, which is one of the main claims in the paper,
    original, or subsumed by [71]?"

We think that the main contribution in Section 6 is to show that an abstract correspondence (expressed indeed by Theorem 24) known in the categorical literature can be extended to account for the relevant structure for interpreting the differential lambda-calculus. Hence, we see Theorems 27 and 29 as the main claims of this section, rather than Theorem 24 (whose statement and proof can be deduced from the results of [71]). We will clarify this.

### 2023-03-18 rebuttal.txt

> "In complex analysis, Laurent series include terms of negative
    degrees, while here I think we only consider terms with
    non-negative degrees."

This is true, yet often in the literature on tropical mathematics (to which we refer, e.g. [59]) these are defined only with positive coefficients. We will explicitly mention it.

> "Is Theorem 24, which is one of the main claims in the paper,
    original, or subsumed by [71]?"

We think that the main contribution in Section 6 is to show that an abstract correspondence (expressed indeed by Theorem 24) known in the categorical literature can be extended to account for the relevant structure for interpreting the differential lambda-calculus. Hence, we see Theorems 27 and 29 as the main claims of this section, rather than Theorem 24 (which can be deduced from the results of [71]). We will clarify this.

### 2023-05-12 TLLA23/BarbaPisto.tex
Commit

\usepackage{macros}
\usepackage{tikz}
  \usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage{stmaryrd,amssymb,proof,cmll,mathrsfs,amsmath,amsthm}
\usepackage{todonotes}

\title{Tropical methods in Lambda-Calculus}
\author{Davide Barbarossa
\institute{DISI, Universit\`a di Bologna}
\email{davide.barbarossa@unibo.it}
\and
\qquad\qquad Paolo Pistone
\institute{\qquad\qquad\qquad DISI, Universit\`a di Bologna}
\email{\qquad\qquad\qquad paolo.pistone@unibo.it}
}
\def\titlerunning{Trends in Linear Logic and Applications}
\def\authorrunning{D.\ Barbarossa and P.\ Pistone}

@@

\begin{abstract}
We propose an interpretation of the $\lambda$-calculus in the framework of tropical mathematics, as a unified framework for both program metrics, based on the analysis of program sensitivity via Lipschitz-conditions, and for resource analysis, based on higher-order program differentiation.
We sketch the relation of this semantics to quantitative properties like differential privacy, convergence logprobabilities and Probabilistic Coherent spaces.
Finally, we study the abstract correspondence between this tropical semantics and Lawvere’s generalised metric spaces.
\end{abstract}

\section*{Motivation}

@@

@@

@@

\section*{Bibliography}

@@

\nocite{*}
\bibliographystyle{eptcs}
%\bibliography{generic}
\end{document}

>>>>

\usepackage{tikz}
  \usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage{stmaryrd,amssymb,proof,cmll,mathrsfs,amsmath,amsthm}
\usepackage{todonotes}

\input{macros}

\title{Tropical methods in Lambda-Calculus?}
\author{Davide Barbarossa
\institute{DISI, Universit\`a di Bologna}
\email{davide.barbarossa@unibo.it}
\and
\qquad\qquad Paolo Pistone
\institute{\qquad\qquad\qquad DISI, Universit\`a di Bologna}
\email{\qquad\qquad\qquad paolo.pistone@unibo.it}
}
\def\titlerunning{Trends in Linear Logic and Applications}
\def\authorrunning{D.\ Barbarossa and P.\ Pistone}

@@

%\begin{abstract}We propose to study the interpretation of the $\lambda$-calculus in the framework of tropical mathematics, as a unified framework for both program metrics -- based on the analysis of program sensitivity via Lipschitz-conditions -- and for resource analysis -- based on higher-order program differentiation.We sketch the relation of this semantics to quantitative properties like differential privacy, convergence logprobabilities and Probabilistic Coherent spaces.Finally, we study the abstract correspondence between this tropical semantics and Lawvere’s generalised metric spaces.\end{abstract}

%\section*{Motivation}

Among the quantitative approaches towards which the programming language community has directed its interest in recent years%(e.g.\ \cite{decarvalho2018, Accattoli2022}, \cite{Breuvart2018, PistoneLICS2022}, \cite{Barthe_2012} \cite{Reed2010, dallago}, \cite{Ehrhard2011, Staton2017}, \cite{difflambda}),
, two different directions have received considerable attention. 
On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask  whether they compute functions which do not differ \emph{too much}, and thus study denotational frameworks in which types are endowed with a metric \cite{Reed2010},\cite{Bonchi2018}, \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
On the other hand, there is the approach based on \emph{differential} \cite{difflambda}, \cite{difflambda}, \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.  or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.

In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.

At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz.
The fundamental motivation of this work is then the observation that 
this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics where, for instance, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.
Tropical mathematics \cite{Simon} is a well established algebraic and geometrical framework, with tight connections with optimisation theory \cite{Sturmfelds}, where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
For instance, the polynomial $p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
$
\varphi(x,y)=\min\{2x, x+2y, 3y\}
$.

Coming back to our discussion on program semantics, tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
Actually, a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.

The viewpoint that we develop in the following sections, may suggests the application of optimization methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions, but it scales to a 
more abstract level, leading to introduce a 
differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).

In this contribution we only sketch some of the main results of our still work in progress analysis.

\section{Tropical mathematics in a nutshell}

More rigorously, we let the \emph{tropical semiring}, the structure at the heart of tropical mathematics, be $([0,\infty], \min, +)$.
Remark that this coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014} (the order is the reverse order $\geq$ on $\BB R$ and the monoid action is addition), the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}, and we will take this point of view in the last section of this contributions.
A tropical polynomial is a piece-wise linear function $f:\Lawv\to \Lawv$ of the form $f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}$
where the $i_{j}$ are natural numbers and the $c_{i_{j}}$ are in $\Lawv$.
For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$
are illustrated in Fig.~\ref{fig:plot1} for $n\leq 4$.
A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes). For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
A \emph{tropical Laurent series} (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function that can be expressed as $f(x)=\inf_{n}\{nx+\matr f_{n}\}$, with $\matr  f_{n}$ a sequence in $\Lawv$. In other words, a tLs is a ``limit'' of tropical polynomials of higher and higher degree. For instance, the function $\varphi(x):=\inf_{n\in\N}\set{nx+\frac{1}{2^n}}$ (illustrated in Fig.~\ref{fig:plot1}), that we will take as our running example, 
is the ``limit'' of the polynomials $\varphi_{n}$. Since $\inf$s are not in general $\min$s, the behavior of tLS may be less predictable than that of tropical polynomials. For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points.

\section{Tropical weighted semantics in a nutshell}

The study of matrices with values over the tropical semiring can be seen as a special case of the
\emph{quantitative relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus and linear logic:
for a fixed \emph{continuous} semi-ring $Q$ [Def. II.5, \cite{Manzo2013}], the category $\QREL$ has sets as objects and set-indexed matrices with coefficients in $Q$ as morphisms, i.e.~$\QREL(X,Y)=Q^{X\times Y}$ (\cite{Manzo2013} would call it $Q^\Pi$).
As it is expected, $Q^X$ is a $Q$-semimodule and the bijection $\hat{(\cdot)}$ identifies the set of linear maps from $Q^X$ to $Q^Y$ with 
 $\HOM{\QREL}{X}{Y}$.

The tropical relational model is thus provided by the category $\LREL$ of matrices with values over $\Lawv$ (which, being a quantale, is indeed a continuous semi-ring).
%, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
%The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
%This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
It is worth observing that the formula for composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
reads as \ $(st)_{a,c}:=\inf\limits_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
 similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$ induced by matrices, which we call \emph{tropical linear}, are exactly those of shape $f(x)_b=\inf\limits_{a\in X} \set{\matr f_{a,b}+x_a}$, for some matrix $\matr f$ from $X$ to $Y$.
%\end{remark}
By applying more or less well-known results (taken from \cite{Manzo2013}, CITARE LEMAYYYY), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:

\begin{fact}
 $\LREL$ is a SMCC, thus a model of the linear $\STLC$.

 The coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.

The exponential $!$ can be ``decomposed'' into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $\LREL$, together with $(!_n)_{n\in\N}$, is a model for $\BSTLC$.

 $\LREL_!$ equipped with $D$, is a CC$\partial$C and the interpretation of the Taylor expansion of a $\STLC$-term $M$, given in \eqref{eq:taylor}, converges to the one of $M$.
\end{fact}

In the previous result, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ is:
$
 \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
 \left\{s_{[b_1,\dots,b_n],c} + \sum\limits_{i=1}^n t_{\mu_i,b_i}\right\}$.
and the \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).

In particular, the Taylor formula decomposes an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as interpreting  a $\lambda$-term 
as a limit of Lipschitz maps, in some sense, thus bridging the metric and differential approaches.  
Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.

Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions? This way, \eqref{eq:taylor} could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps. In other words, unbounded term application could be seen 
as a limit of \emph{more and more sensitive} operations.

@@

 As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a linear map $\Lawv^{!X}\to\Lawv^Y$, but we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting 
 $t^!(x):=t\circ_! x$.
%  (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
 This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
 Concretely, we have
 \begin{equation}
 t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}
 \end{equation}
 where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
 These functions correspond then to tLs with possibly infinitely many variables (in fact, as many as the elements of $X$).

Therefore, we study in this section the properties of those functions, from the point of view of mathematical analysis.
%In the following we will also refer to them as tLs.

% 
% We will call them simply \emph{tropical Laurent series (tLs)}.
% %Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
%\end{remark}
%
%We find the usual notion of tLs of one variable as follows:

%\begin{remark}
Notice that, by identifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tLs generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{n\in\N}\set{nx+\matr f(n)}$, for some $\matr f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
In a similar way, the tropical polynomials can be identified with the tLs
$f:\Lawv^X\to\Lawv^Y$ 
 for which the support $\C F=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ is \emph{finite}, and which have thus shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$.
 This is again the generalisation of usual tropical polynomials to the case of infinitely many variables.

@@

We already remarked that the tropical semiring $\Lawv$ coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$.
In particular, a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we explicitly take in Section \ref{section6}).

\begin{remark}
 The category $\QREL$ is (equivalent to) a subcategory of the category $Q\SF{Mod}$ of \emph{complete} $Q$-semimodules.
 If $\QREL$ corresponds to considering semimodules (the $Q^X$'s) whose vectors are given in coordinates w.r.t.\ a \emph{fixed base} (the set $X$), $Q\SF{Mod}$ corresponds to considering semimodules in abstract, without fixing a base.
 We take this viewpoint in Section~\ref{section6}.
\end{remark}

@@

%\section*{Bibliography}

@@

%\nocite{*}
%\bibliographystyle{eptcs}
%\bibliography{generic}

%\bibliographystyle{plain}
%\bibliography{tropical.bib}

\end{document}
